---
title: "SALAZAR_gis"
author: "Jae Salazar"
date: "2026-06-22"
output: html_document
---
## Assigned species: Northern Sierra Madre forest monitor (*Varanus bitatawa*) ##

## Load all libraries. ##
```{r}
library(maps)
library(osmdata)
library(sf)
library(terra)
library(ggplot2)
library(dplyr)
library(ggnewscale)
```

## Clean the data by removing records found outside the reported species range. ##
```{r}
varanus <- read.csv("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 10/Varanus bitatawa.csv")
varanusclean <- varanus[complete.cases(varanus[, c("decimalLatitude", "decimalLongitude")])]
```

## From the Philippine administrative boundaries/areas, get only the provincial level, and subset the provinces that contain only distribution records for your assigned species. ##
```{r}
provinces <- st_read("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 10/PHL_adm/PHL_adm1.shp")

provset <- provinces %>%
  filter(NAME_1 %in% c("Cagayan",
                      "Isabela",
                      "Quirino",
                      "Aurora"))
```

## Download the Philippine digital elevation model, then clip it using the provinces you selected in step 5.
```{r}
phil_dem <- rast("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 10/ph_ strm_geo.tif")
print(st_crs(provset) != st_crs(phil_dem))
provset <- st_transform(provset, st_crs(phil_dem))
dem <- crop(phil_dem, provset) |> mask(provset)
names(dem) <- "elevation"
r_df <- as.data.frame(dem, xy = TRUE)
```

## Now, put all the data together into a map, following the pre-lab lecture. Make sure that data layers (points, polygons, DEMs) are arranged properly.##

```{r, fig.width=8, fig.height=12}
ggplot() +
  geom_tile(data = r_df, aes(x = x, y = y, fill = elevation)) +
    scale_fill_gradientn(
    colors = c("#2b83ba", "#abdda4", "#ffffbf", "#fdae61", "#d7191c"),
    name = "Elevation (m)") +
    new_scale_fill() +
  geom_point(data = varanusclean,
    aes(x = decimalLongitude, y = decimalLatitude),
    shape = 21,
    fill = "black",
    size = 3,
    stroke = 1.2) +
  coord_equal() +
  labs(title = "Species Occurrence Area",
    x = "Longitude",
    y = "Latitude") +
  theme_bw()
```
