---
title: "Activity 7: Clustering (K-means)"
author: "Jae Salazar"
date: "2026-06-08"
output: html_document
---

# K-means (single K)
## Step 2: Read in the morphology data as a dataframe.
```{r}
morphodata <- read.csv("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 7/morphology_data.csv")
```

## Step 3: Remove the sex column.
```{r}
morphonew <- subset(morphodata, select = -c(Sex, Pop))
```

## Step 4: Ensure that the dataframe has no NAs.
```{r}
morphoclean <- morphonew[complete.cases(morphonew), ]
```

## Step 5: Prior to performing K-means, standardize all values using the “scale” function.
```{r}
morphoscaled <- scale(morphoclean)
```

## Step 6: Perform K-means with 3 cluster centers using 25 initial random centroids.
```{r}
kmeans1 <- kmeans(morphoscaled, centers = 3, nstart = 25)
```

## Step 7: Show the cluster assignment and a summary of results by simply printing them to the console.
```{r}
head(kmeans1$cluster)
kmeans1
```

## Step 8: Create a scatter plot with points colored by the assigned cluster.
```{r, fig.width=6, fig.height=4}
plot (
  morphoscaled[, 1:2],
  col = kmeans1$cluster,
  main = "K-means Clustering of Philippine Green Crested Lizards Data"
  )
```

# K-means (loop)
## Step 1: Now, create a loop function to perform K-means with cluster centers from 2 to 10.
```{r}
ss <- numeric(10)

for(k in 1:10) {
  kmeans2 <- kmeans(morphoscaled, centers = k, nstart = 25)
  ss[k] <- kmeans2$tot.withinss
}
```

## Step 2: From the output of your loop function, create an elbow curve plot to determine the optimal K.
```{r, fig.width=6, fig.height=4}
library(factoextra)
fviz_nbclust(
  morphoscaled,
  kmeans,
  method = "wss",
  k.max = 10
  )
```
```{r, fig.width=6, fig.height=4}
plot(1:10, ss, type = "b", pch = 19, frame = FALSE, 
        xlab="Number of clusters k",
        ylab="Total Within Sum of Square",
        main = "Optimal Number of Clusters")
```

## Step 3: Recreate steps 6–8 above using the optimal K determined by the elbow curve.
## 3a
```{r}
kmeans3 <- kmeans(morphoscaled, centers = 2, nstart = 25)
```
## 3b
```{r}
head(kmeans3$cluster)
kmeans3
```
## 3c
```{r, fig.width=6, fig.height=4}
plot (
  morphoscaled[, 1:2],
  col = kmeans3$cluster,
  main = "K-means Re-clustering of Philippine Green Crested Lizards Data"
  )
```