---
title: "Activity 3: Summary statistics, frequency distribution, and basic graphical representation"
author: "Jae Salazar"
date: "01 June 2026"
output: html_document
---
# *Descriptive statistics*
## **Step 1** | Read in the CSV file as a dataframe using the function “read.csv”. Make sure to clean the data prior to performing the tasks below. Do not attempt to clean the data in Excel. Perform all cleaning (e.g., NA removal) in R using the code we discussed.
```{r}
herpedata <- data.frame(read.csv("D:/Files//DLSU Files/Year 3 Term 3/LBBBI27/Activity 3/herpetozoa-033-095-s001.csv"))
cleaningdata <- complete.cases(herpedata)
herpeclean <- herpedata[cleaningdata, ]
herpeclean
```

## **Step 2** | Show the data structure.
```{r}
class(herpeclean)
```

## **Step 3** | Subset the elevation data and provide the mean, standard deviation, and range.
```{r}
elevation <- herpeclean["Elev"]
elevationdata <- c(sapply(elevation, mean), sapply(elevation, sd), sapply(elevation, range))
elevationdata
```

## **Step 4** | Subset the species data (columns G to AG) and count the total number of observations per species. Show two methods, using a typical loop and a compact loop.
```{r}
names(herpeclean)
speciesdata <- herpeclean[, 7:33]
```
## *Typical loop*
```{r}
## Placeholder to create the speciestotals object for use in the typical loop
speciestotals <- 0
for(i in seq_len(ncol(speciesdata))) {
    speciestotals[i] <- print(sum(speciesdata[i]))
 }
speciestotal <- matrix(speciestotals, nrow = 27, ncol = 1)
dimnames(speciestotal) <- list(names(speciesdata), "Total")
speciestotal
```

## *Compact loop*
```{r}
speciestotal <- matrix(sapply(speciesdata, sum))
dimnames(speciestotal) <- list(names(speciesdata), "Total")
speciestotal
## Phlo has the highest (377), Babu, Geat, Hety, Bosc, Drru, Apbo have the lowest (1)
```

## **Step 5** | Combine the columns of species with the most and fewest observations with habitat types (Htypes). Determine the mean count of these species in each habitat type.
```{r}
specieshabitats <- herpeclean[c("Htypes", "Phlo", "Babu", "Geat", "Hety", "Bosc", "Drru", "Apbo")]
specieshabitats

## Subset into habitats
ASG <- specieshabitats[specieshabitats$Htypes == "ASG", -1]
ESG <- specieshabitats[specieshabitats$Htypes == "ESG", -1]
CVT <- specieshabitats[specieshabitats$Htypes == "CVT", -1]
OGF <- specieshabitats[specieshabitats$Htypes == "OGF", -1]

## Generate means per habitats
ASGdata <- sapply(ASG, mean)
ESGdata <- sapply(ESG, mean)
CVTdata <- sapply(CVT, mean)
OGFdata <- sapply(OGF, mean)

## Combine the means into one matrix
habitatdata <- cbind(ASGdata, CVTdata, ESGdata, OGFdata)
habitatdata

## OGF has the highest species counts, ESG has none.
```
# *Graphical representation*
## **Step 1** | From the original dataframe, combine the species total count column (SPPTotal) with the elevation and habitat type columns and create a scatter plot where elevation should be on the x-axis, while species total count is on the y-axis. Change the point colors based on habitat type.
```{r scat, fig.width = 4, fig.height = 3}
plotdata <- herpeclean[4:6]
colors <- c(ASG = "red", CVT = "blue", ESG = "green", OGF = "yellow")
## Generate scatterplot
plot(plotdata$Elev, plotdata$SPPTotal, col = colors[plotdata$Htypes], main = "Species Totals per Elevation", xlab = "Elevation", ylab = "Species Totals")
```

## **Step 2** | From the combined data of species total counts, elevation, and habitat types, determine the total/sum number of species observed per habitat, and create a bar plot (use the bar plot help) where the x-axis shows the habitat types and the y-axis shows the total species.
```{r bar, fig.width = 4, fig.height = 3}
ASGtotal <- plotdata[plotdata$Htypes == "ASG", -(1:2)]
ESGtotal <- plotdata[plotdata$Htypes == "ESG", -(1:2)]
CVTtotal <- plotdata[plotdata$Htypes == "CVT", -(1:2)]
OGFtotal <- plotdata[plotdata$Htypes == "OGF", -(1:2)]
## Combine sums of the subsets for barplot data
bardata <- c(sum(ASGtotal), sum(ESGtotal), sum(CVTtotal), sum(OGFtotal))
## Generate barplot
barplot(bardata, main = "Species Totals per Habitat Type", xlab = "Habitat Types", ylab = "Species totals", names.arg = c("ASG", "ESG", "CVT", "OGF"))
```

## **Step 3** | Lastly, create a histogram of total species counts using the histogram function. Make sure all axes are properly labeled.
```{r hist, fig.width = 4, fig.height = 3}
hist(plotdata$SPPTotal, main = "Total Species Counts", xlab = "Total Species Counts", ylab = "Frequency")
```

