---
title: "Activity 11 - Linear Regression"
author: "Jae Salazar"
date: "2026-07-05"
output: html_document
---
## Load all libraries. ##
```{r}
library(sf)
library(terra)
library(dplyr)
```

# Simple linear regression #

## 2. Read in the species data and remove unnecessary rows with NAs. ##
```{r}
herpedata <- data.frame(read.csv("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 3/herpetozoa-033-095-s001.csv"))
cleaningdata <- complete.cases(herpedata)
herpeclean <- herpedata[cleaningdata, ]
```

## 3. Read in the elevation data (ph_strm_geo.tif). ##
```{r}
phil_dem <- rast("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 10/ph_ strm_geo.tif")
```

## 4. From the species data, subset the longitude and latitude coordinates. Then convert the coordinates to a spatial vector using the “vect” function from the “terra” package. ##
```{r}
herpecoord <- herpeclean[2:3]
coord <- vect(herpecoord)
```

## 5. Using the converted coordinates, extract elevation data for each point with the “extract” function in the “terra” package. The extracted data will have two columns: “ID” and “Band_1.” The “Band_1” column contains the elevation. ##
```{r}
coordelev <- extract(phil_dem, coord)
```

## 6. Subset the species data to retain only the species richness column (SPPTotal). Then combine this column with the extracted elevation. Rename “Band_1” to “elevation.” ##
```{r}
richness <- herpeclean$SPPTotal
richelev <- cbind(richness, coordelev$Band_1)
colnames(richelev) <- c("Richness", "Elevation")
richelev <- as.data.frame(richelev)
richelev
```

## 7. Now, perform simple linear regression using the code from the pre-lab lecture, including a plot. What can you infer about the relationship between species and elevation? ##
```{r, fig.width = 10, fig.height = 5}
x <- richelev$Elevation
y <- richelev$Richness
xbar <- mean(x)
ybar <- mean(y)
b <- sum((x - xbar) * (y - ybar)) / sum((x - xbar)^2)
a <- ybar - b * xbar
cat("Regression Equation:\n")
cat("y=", round(a,3), "+", round(b,3), "x\n")
yhat <- a + b * x
plot(x, y,
  pch = 19,
  col = "forestgreen",
  xlab = "Elevation (m)",
  ylab = "Species Richness",
  main = "Simple Linear Regression Model")
abline(a, b, col = "goldenrod4", lwd = 2)
points(x, yhat, pch = 17, col = "mediumblue", cex = 1.2)
segments (x0 = x, y0 = y, x1 = x, y1 = yhat, lty = 2, col = "darkorange")
legend("topright",
  legend = c("Observed", "Predicted", "Regression Line"),
  pch = c(19, 17, NA),
  lty = c(NA, NA, 1),
  col = c("forestgreen", "mediumblue",  "goldenrod4"),
  lwd = c(NA, NA, 2),
  bty = "n")
```

While the observed values seem to be more spread out across elevations, the best fitted line predicts that species richness increases as elevation decreases.

# B. Multiple linear regression #
## 1. Read in both annual precipitation and temperature data and extract the data following the instructions from A4–5. ##
```{r}
phil_rain <- rast("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 11/annual_precipitation.tif")
phil_temp <- rast("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 11/annual_temperature.tif")
coordrain <- extract(phil_rain, coord)
coordtemp <- extract(phil_temp, coord)
```

## 2. Combine the extracted data with the species and elevation data into a single dataframe. ##
```{r}
combined <- cbind(richness, coordelev$Band_1, coordrain$annual_precipitation, coordtemp$annual_temperature)
colnames(combined) <- c("Richness", "Elevation", "Annual Precipitation", "Annual Temperature")
combined <- as.data.frame(combined)
```

## 3. Perform multiple linear regression using the “lm” function.
```{r}
x <- combined$Elevation
x1 <- combined$`Annual Precipitation`
x2 <- combined$`Annual Temperature`
y <- combined$Richness
model <- lm(y ~ x + x1 + x2)
coef(model)
a <- coef(model)[1]
b <- coef(model)[2]
c <- coef(model)[3]
d <- coef(model)[4]
cat("Regression Equation:\n")
cat("y=", round(a,3), "+", round(b,3), "x", "+", round(c,3), "x1", "+", round(d,3), "x2\n\n")
yhat <- predict(model)
residuals <- residuals(model)
results <- data.frame(
  Elevation = x,
  "Annual Precipitation" = x1,
  "Annual Temperature" = x2,
  Observed = y,
  Predicted = round(yhat,2),
  Residual = round(residuals,2)
)
print(results)
summary(model)
```
The computed model shows that **the annual precipitation and annual temperature readings had significance levels of less than 0.05.** Thus, **these variables showed statistical significance in predicting the levels of species richness.**