---
title: "Random Forest regression"
author: "Jae Salazar"
date: "2026-07-13"
output: html_document
---

## 2. Read in the acoustic data and remove unnecessary columns (i.e., column 1). Remember that your response variable is the species column. ##
```{r}
acousticdata <- data.frame(read.csv("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 13/call_data_12October2025.csv"))
acousticlean <- acousticdata[complete.cases(acousticdata), ]
acousticlean$File_Name <- NULL
acousticlean$Species <- as.factor(acousticlean$Species)
```

## 3. Design your Cartesian grid and perform a random forest that iterates on each hyperparameter combination. ##
```{r}
library(randomForest)
library(ranger)

mtry <- floor(sqrt(ncol(acousticlean) -1))

grid <- expand.grid(
  num.trees = c(100, 200, 300, 400, 500),
  mtry = 3,
  min.node.size = c(1, 2, 3, 4, 5)
)

results <- data.frame(
  num.trees = numeric(),
  mtry = numeric(),
  min.node.size = numeric(),
  accuracy = numeric(),
  oob.error = numeric()
)

for(i in 1:nrow(grid)){
  model <- ranger(
    Species ~ .,
    data = acousticlean,
    num.trees = grid$num.trees[i],
    mtry = grid$mtry[i],
    min.node.size = grid$min.node.size[i]
  )
     
  pred <- predict(model, data = acousticlean)$predictions
  acc <- mean(pred == acousticlean$Species)
  results <- rbind(
    results,
    cbind(grid[i,], accuracy = acc, oob.error = model$prediction.error)
  )
}
```

## 4. Pick the best hyperparameter combination that minimizes the out-of-bag error. ##
```{r}
results[which.min(results$oob.error), ]
```

## 5. Using the optimal hyperparameter, provide a plot of variable importance and the predicted class probabilities. ##
```{r, fig.width = 8, fig.height = 8}
library(rpart)
library(rpart.plot)

response <- "Species"
set.seed(123)
ntree <- 100
rf_model <- randomForest(
  Species ~ .,          
  data = acousticlean,
  ntree = ntree,
  mtry = mtry,
  importance = TRUE
)

varImpPlot(rf_model)

rpart.plot(rpart(Species ~ ., data = acousticlean))
```
