---
title: "SALAZAR_logistic_reg"
author: "Jae Salazar"
date: "2026-07-09"
output: html_document
---
## Load libraries. ##
```{r}
library(maps)
library(osmdata)
library(sf)
library(terra)
library(ggplot2)
library(dplyr)
library(ggnewscale)
```

## 2. Read in the species data and remove unnecessary rows with NAs. ##
```{r}
mousedeerdata <- data.frame(read.csv("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 12/ph_mouse_deer.csv"))
mousedeerclean <- mousedeerdata[complete.cases(mousedeerdata), ]
```

## 3. Generate a map showing the species' observation localities using the longitude and latitude coordinates. Check Activity 10 and use the provided codes as your guide for creating the map. Export the map as a PDF file. ##
```{r, fig.width = 8, fig.height = 8}
provinces <- st_read("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 10/PHL_adm/PHL_adm1.shp")
palawan <- provinces %>%
  filter(NAME_1 %in% c("Palawan"))
phil_dem <- rast("D:/Files/DLSU Files/Year 3 Term 3/LBBBI27/Activity 10/ph_ strm_geo.tif")
print(st_crs(palawan) != st_crs(phil_dem))
palawan <- st_transform(palawan, st_crs(phil_dem))
dem <- crop(phil_dem, palawan) |> mask(palawan)
names(dem) <- "elevation"
r_df <- as.data.frame(dem, xy = TRUE)
mousedeercoor <- mousedeerclean[, c("Longitude", "Latitude")]
ggplot() +
    geom_tile(data = r_df, aes(x = x, y = y, fill = elevation)) +
    scale_fill_gradientn(
        colors = c("#2b83ba", "#abdda4", "#ffffbf", "#fdae61", "#d7191c"),
        name = "Elevation (m)") +
    new_scale_fill() +
    geom_point(data = mousedeercoor,
               aes(x = Longitude, y = Latitude),
               shape = 21,
               fill = "black",
               size = 2,
               stroke = 0.5) +
    coord_equal() +
    labs(title = "Species Occurrence Area",
         x = "Longitude",
         y = "Latitude") +
    theme_bw()
```

## 4. Using the “glm()” function, perform multiple logistic regression. Ensure that the habitat types (i.e., HabitatCode) are treated as categorical variables by converting them to factors. ##
```{r}
mousedeerlm <- mousedeerclean[, c("AverageDistance", "Effort", "HabitatCode", "Detected")]
mousedeerlm$HabitatCode <- as.factor(mousedeerlm$HabitatCode)
x1 <- mousedeerlm$AverageDistance
x2 <- mousedeerlm$Effort
x3 <- mousedeerlm$HabitatCode
y <- mousedeerlm$Detected
data <- data.frame(x1, x2, x3, y)
model <- glm(y ~ x1 + x2 + x3,
  data = data,
  family = binomial)
data$Probability <- predict(model, type = "response")
data$Prediction <- ifelse(data$Probability >= 0.5, 1, 0)
print(data)
table(Observed = data$y, Predicted = data$Prediction)
```

## 5. After fitting the model, create a logistic regression curve plot for each independent variable used. Export the plots as PDF files. ##

Average Distance vs Probability of Detection
```{r, fig.width = 8, fig.height = 8}
new_x1 <- data.frame(
  x1 = seq(min(data$x1), max(data$x1), length.out = 19),
  x2 = mean(data$x2),
  x3 = levels(data$x3)[1]
)
new_x1$prob <- predict(model, newdata = new_x1, type = "response")
ggplot()+
  geom_jitter(data = data,
    aes(x = x1, y = y),
    height = 0.03,
    width = 0,
    alpha = 0.05,
    color = "palegreen3") +
  geom_line(data = new_x1,
    aes(x = x1, y = prob),
    color = "royalblue",
    linewidth = 1.2)+
  geom_point(data = data,
    aes(x = x1, y = Probability),
    color = "red",
    alpha = 0.6,
    size = 2) +
  labs(x = "Average Distance",
    y = "Probability of Detection")+
  theme_bw()
```

Effort v Probability of Detection
```{r, fig.width = 8, fig.height = 8}
new_x2 <- data.frame(
x1 = mean(data$x1),
x2 = seq(min(data$x2), max(data$x2), length.out = 19),
x3 = levels(data$x3)[1]
)
new_x2$prob <- predict(model, newdata = new_x2, type = "response")

ggplot()+
  geom_jitter(data = data,
    aes(x = x2, y = y),
    height = 0.03,
    width = 0,
    alpha = 0.05,
    color = "palegreen3") +
  geom_line(data = new_x2,
    aes(x = x2, y = prob),
    color = "royalblue",
    linewidth = 1.2)+
  geom_point(data = data,
    aes(x = x2, y = Probability),
    color = "red",
    alpha = 0.6,
    size = 2) +
  labs(x = "Effort",
    y = "Probability of Detection")+
  theme_bw()
```

Habitat Type v Probability of Detection
```{r, fig.width = 8, fig.height = 8}
new_x3 <- expand.grid(
  x1 = mean(data$x1),
  x2 = mean(data$x2),
  x3 = levels(data$x3)
)
new_x3$prob <- predict(model, newdata = new_x3, type = "response")
ggplot(new_x3,
    aes(x = x3, y = prob)) +
  geom_point(size = 4, colour = "blue") +
    ylim(0,1) +
  labs(x = "Habitat",
    y = "Predicted Probability") +
  theme_bw()
```

## 6. Now, output the model summary and examine the results. ##
```{r}
summary(model)
```

